Neural Models for Target-Based Computer-Assisted Musical Orchestration: A Preliminary Study

### 2021-06-24 csmc_mume_2020/paper/JCMS_2021/paper.tex
Revisions from reviews 1 and 2

In this paper, we try a different approach to this problem by experimenting with deep neural architectures. The main idea is to train a model to classify combinations of real instruments and then use it for orchestration. A typical solution for assisted orchestration is a set of triples \emph{instrument-pitch-dynamics} such as \{\texttt{Flute C6 pp, Bassoon C4 mf, Bassoon G4 ff}\}. By training a neural network with real combinations of instrumental notes, it will acquire the ability to identify the presence of each instrument and its associated pitch by building the appropriate latent representation. Thus, when an unknown target sound is given as input, the network will identify which are the best instruments to match the target sound, and it will be able to  deconstruct a complex mixture of timbres into individual instrument notes. This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07, Kitahara05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16, Bian19}.

@@

@@

In order to get a better sense on the complexity of the problem, we tested three baseline classifiers: support vector machine (SVM), random forest (RF), and K-nearest neighbors (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15.  All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length.

In this case, differently from the neural models, the features used are the MFCCs of the resulting combination; we chose to use MFCCs for this setting, instead of the Mel spectrogram, to have a lower number of features, that is more manageable by parametric classifiers. We found SVM to have the highest accuracy of the three classifiers across all experiments.

@@

Following the first three convolutional layers, there is an LSTM layer which outputs 32 matrices. After the LSTM layer, there is a final convolutional layer yielding a tensor of dimensions $32\times 8 \times 21$. We flatten the outputs and feed them into a fully connected layer with Dropout, then another another fully connected layer. Finally, the sigmoid function is applied to the final layer. Since each class is independent, we are able to take the sigmoid activation and use binary classification for each class. The model architecture is shown in Figure \ref{cnn_lstm}.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_CNN.png}
    \caption{CNN with LSTM}
    \label{best_acc_cnn}
  \end{subfigure}%
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_ResNet.png}
    \caption{ResNet}
    \label{best_acc_resnet}
  \end{subfigure}
  \caption{Plot showing per instrument accuracies for CNN with LSTM and ResNet depending on the number of samples present in the combinations}
  \end{figure}

@@

\begin{table*}
	\centering    
  \resizebox{\columnwidth}{!}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}
     \hline
      \textbf{Model} & \textbf{Ob + Bn} & \textbf{Bn} & \textbf{Bass cl.} & \textbf{Bell 1} & \textbf{Bell 2} & \textbf{Multiph. 1} & \textbf{Car horn} & \textbf{Boat} $\ldots$ \\
      \hline
      CNN with LSTM & 0.17 & 0.28 & 0.70 & 0.55 & 0.26 & \textbf{1.10} & 0.68 & \textbf{1.12} $\ldots$  \\
      \hline
      ResNet & 0.34 & 0.50 & 0.48 & 0.59 & 0.45 & 0.90 & 0.49 & \textbf{1.16} $\ldots$ \\
      \hline
      \hline
      \textbf{$\ldots$} & \textbf{Wind harp} & \textbf{Chord 1} & \textbf{Multiph. 2} & \textbf{Chord 2} & \textbf{Gong} & \textbf{Scream} & \textbf{Brass} & \textbf{Average}  \\
      \hline
      $\ldots$ & 0.55 & 0.79 & 0.70 & 0.57 & 0.73 & \textbf{1.14} & 0.79 & 0.71\\
      \hline
      $\ldots$  & 0.61 & 0.86 & 0.51 & 0.37 & 0.71 & \textbf{1.03} & \textbf{1.05} & 0.66 \\
      \hline
    \end{tabular}}
    \caption{Quantitative comparison of orchestrations as ratios to Orchidea. Eqn. \ref{distance} was used to compute distances between orchestrations and targets. What is shown is the ratio between the distance of Orchidea's solution to the target and our solution's distance to the same target. A value less than 1 means that our model performed worse (i.e. had a larger distance), and a value greater than 1 means our model performed better than Orchidea. The last column shows the ratio of the average distances for the model across all targets.}\label{orch_eval}
\end{table*}

@@

In order to have a better comparison, we did not allow Orchidea to use any of the advanced features: we did not apply any symbolic constraints or harmonic analysis and we forced it forced to use all 10 instruments in each solution. This creates a more fair comparison, since our models are unable to create constrained or sparse solutions.

Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that loosely represented a C minor chord, so we checked if the orchestration contained the notes of the chord. If a target had specific notes, we identified if the note or its partials were present. For example there is target of an Oboe playing an A4 and a Bassoon playing a C\#3, the solution from ResNet contained the partials of the Bassoon's note: C\#3, G\#4, C\#5, and E6 were all included on various instruments in the solution.

@@

We calculated the distance between the target samples and our orchestrated solutions. We then orchestrated the same targets with Orchidea and calculated the distance for Orchidea's solutions. A comparison of these results is in Table \ref{orch_eval}. While our model is not able to outperform Orchidea, it shows consistent results. We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target.

>>>>

In this paper, we try a different approach to this problem by experimenting with deep neural architectures. Unlike Orchidea, which uses an evolutionary optimization algorithm to find a solution, we frame computer-assisted orchestration as a classification task. The main idea is to train a model to classify combinations of real instruments and then use this model for orchestration. Where Orchidea has multiple pre- and post-processing steps, we create an end-to-end network that takes a target sound as input and outputs a list of samples that can be combined to create the orchestrated solution. A typical solution for assisted orchestration is a set of triples \emph{instrument-pitch-dynamics} such as \{\texttt{Flute C6 pp, Bassoon C4 mf, Bassoon G4 ff}\}. By training a neural network with real combinations of instrumental notes, it will acquire the ability to identify the presence of each instrument and its associated pitch by building the appropriate latent representation. Thus, when an unknown target sound is given as input, the network will identify which are the best instruments to match the target sound, and it will be able to  deconstruct a complex mixture of timbres into individual instrument notes. This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07, Kitahara05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16, Bian19}.

@@

We chose to perform data augmentation solely for increasing the robustness of the models, not for any musical reasons. Data augmentation is only applied during training, so no pitch shifting or feature dropout occurs when orchestrating a target.

@@

In order to get a better sense of the complexity of the problem, we tested three baseline classifiers: support vector machine (SVM), random forest (RF), and K-nearest neighbors (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15.  All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length.

In this case, differently from the neural models, the features used are the Mel-frequency cepstral coefficients (MFCCs) of the resulting combination. We chose to use MFCCs for this setting, instead of the Mel spectrogram, to have a lower number of features which is more manageable for parametric classifiers. We found SVM to have the highest accuracy of the three classifiers across all experiments.

@@

Following the first three convolutional layers, there is an LSTM layer which outputs 32 matrices. After the LSTM layer, there is a final convolutional layer yielding a tensor of dimensions $32\times 8 \times 21$. We flatten the outputs and feed them into a fully connected layer with Dropout, then another fully connected layer. Finally, the sigmoid function is applied to the final layer. Since each class is independent, we are able to take the sigmoid activation and use binary classification for each class. The model architecture is shown in Figure \ref{cnn_lstm}.

@@

\begin{figure}[h]
  \centering
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_CNN.png}
    \caption{CNN with LSTM}
    \label{best_acc_cnn}
  \end{subfigure}%
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_ResNet.png}
    \caption{ResNet}
    \label{best_acc_resnet}
  \end{subfigure}
  \caption{Plot showing per instrument accuracies for CNN with LSTM and ResNet depending on the number of samples present in the combinations}
  \end{figure}

@@

\begin{table*}
	\centering    
  \resizebox{\columnwidth}{!}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}
     \hline
      \textbf{Model} & \textbf{Ob + Bn} & \textbf{Bn} & \textbf{Bass cl.} & \textbf{Bell 1} & \textbf{Bell 2} & \textbf{Multiph. 1} & \textbf{Car horn} & \textbf{Boat} \\
      \hline
      CNN with LSTM & 0.17 & 0.28 & 0.70 & 0.55 & 0.26 & \textbf{1.10} & 0.68 & \textbf{1.12} \\
      \hline
      ResNet & 0.34 & 0.50 & 0.48 & 0.59 & 0.45 & 0.90 & 0.49 & \textbf{1.16} \\
      \hline
      \hline
      \textbf{Model} & \textbf{Wind harp} & \textbf{Chord 1} & \textbf{Multiph. 2} & \textbf{Chord 2} & \textbf{Gong} & \textbf{Scream} & \textbf{Brass} & \textbf{Average}  \\
      \hline
      CNN with LSTM & 0.55 & 0.79 & 0.70 & 0.57 & 0.73 & \textbf{1.14} & 0.79 & 0.71\\
      \hline
      ResNet  & 0.61 & 0.86 & 0.51 & 0.37 & 0.71 & \textbf{1.03} & \textbf{1.05} & 0.66 \\
      \hline
    \end{tabular}}
    \caption{Quantitative comparison of orchestrations as ratios to Orchidea. Eqn. \ref{distance} was used to compute distances between orchestrations and targets. What is shown is the ratio between the distance of Orchidea's solution to the target and our solution's distance to the same target. A value less than 1 means that our model performed worse (i.e. had a larger distance), and a value greater than 1 means our model performed better than Orchidea. The last column shows the ratio of the average distances for the model across all targets.}\label{orch_eval}
\end{table*}

@@

In order to have a better comparison, we did not allow Orchidea to use any of the advanced features: we did not apply any symbolic constraints or harmonic analysis and forced it to use all 10 instruments in each solution. This creates a more fair comparison, since our models are unable to create constrained or sparse solutions.

Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example: one of the samples of a bell had partials that loosely represented a C minor chord, so we checked whether the orchestration contained the notes of this chord. If a target included specific notes, we identified whether the note or its partials were present. For example there is a target of an Oboe playing an A4 and a Bassoon playing a C\#3. ResNet's solution for this target contained the partials of the Bassoon's note: C\#3, G\#4, C\#5, and E6 were all included on various instruments in the solution.

@@

We calculated the distance between the target samples and our orchestrated solutions. We then orchestrated the same targets with Orchidea and calculated the distance for Orchidea's solutions. A comparison of these results is in Table \ref{orch_eval}. While our model is not able to outperform Orchidea, it shows consistent results. We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Quantitatively, CNN performs better with an average ratio of $0.71$ compared to ResNet with an average ratio of $0.66$. Both models outperform Orchidea on three out of fifteen of the targets. Through our qualitative inspection, we find that CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target.
